---
title: "Introduction to DPLYR and RMarkdown"
date: '`3/31/2020'
output:
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(warnings = FALSE)
knitr::opts_chunk$set(messages = FALSE)
```
# Today's Files


# RMarkdown 

Before we get started with the lesson on dplyr, I want us to all try and knit together a simple markdown file. To do this, we are going to go to `File -> New File -> R Markdown ->` and then provide a title for our Markdown file. 

From here, we are going to work on creating both an html file and a pdf file from a Markdown document. This can be tricky at times, so if you have issues, please let Marieke or David know after class and we can help you work through them. 

# Data Preparation

# Packages for Today
Today we are going to be using the dplyr package and the readr package. Since they are both contained within the tidyverse package, we will just load the tidyverse package. 

We are also going to be reading in a dataset from the `schrute` package, so we will need to intsall and load up that package.

```{r}
library(tidyverse)
#install.packages("schrute")
library(schrute)
```

Read in Data from the `schrute` package

```{r}
lines <- theoffice
office_ratings <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-17/office_ratings.csv')

```

The lines dataset has every line of dialogue spoken throughout the TV series "The Office". The office_ratings dataset just has information on each episode.

## dplyr

dplyr is a package in R that allows you to work with and manipulate your data. It will allow us to focus in on the variables (columns) of interest and the observations (rows) of interest. 

## The Pipe

The pipe is an operator in R that allows you to chain together functions in dplyr. In the past, to use multiple functions you would have to nest your functions inside of each other, but using the pipe allows you to chain the functions together in a readable/reproducible format. The pipe character is `%>%` and essentially means "then" You can create the character using `CTRL+SHIFT+M` or `CMD+SHIFT+M` 

The pipe essentially takes whatever is before the pipe as input for whatever is after the pipe. 

```{r}
View(lines)

lines %>% 
  View()

office_ratings %>% 
  head(3)

office_ratings %>% 
  head(5) %>% 
  tail(1)
```

## distinct/count

The distinct function will provide for you the distinct values of a column, while count provides you with both the distinct values of a column and then number of times each value shows up. The following example investigates the different characters (`character`) in "The Office" from the lines dataset:

```{r}
lines %>% 
  distinct(character)

#Pipe into nrow() to get the number of distinct characters (other functions also exist to do something similar)
lines %>% 
  distinct(character) %>% 
  nrow()

lines %>% 
  count(character)
```

## Exercise 
Using the `lines` dataframe: 

1. Using the pipe, get the summary of the dataframe

2. How many distinct writers did they have on the show? 

3. How many times does B.J. Novak come up as a director?

<details><summary>Click here for the answers </summary>

```{r}
#1.
lines %>% 
  summary()

#2.
lines %>% 
  distinct(writer) %>% 
  nrow()
  #47
#3. 
lines %>% 
  count(director)
  #1307
```
</details>

-------------------------

## Arrange

The `arrange()` function does what it sounds like. It takes a data frame or tbl and arranges (or sorts) by column(s) of interest. The first argument is the data, and subsequent arguments are columns to sort on. Use the `desc()` function to arrange by descending.

The following code would get the number of lines each character spoke throughout the show:

```{r}
lines %>% 
  count(character) %>% 
  arrange(n)

# Since the default is ascending order, we are not getting the results that are probably useful, so let's use the desc() function
lines %>% 
  count(character) %>% 
  arrange(desc(n))

office_ratings %>% 
  arrange(imdb_rating)
```

## filter

If you want to filter **rows** of the data where some condition is true, use the `filter()` function. 

1. The first argument is the data frame you want to filter, e.g. `filter(mydata, ...`.
2. The second argument is a condition you must satisfy, e.g. `filter(clean, variable == "levelA")`.

- `==`: Equal to
- `!=`: Not equal to
- `>`, `>=`: Greater than, greater than or equal to
- `<`, `<=`: Less than, less than or equal to

If you want to satisfy *all* of multiple conditions, you can use the "and" operator, `&`. 

The "or" operator `|` (the pipe character, usually shift-backslash) will return a subset that meet *any* of the conditions.

```{r}
office_ratings %>% 
  filter(season == 1)

office_ratings %>% 
  filter(imdb_rating >= 9.0)

lines %>% 
  filter(imdb_rating >= 9.0) %>% 
  distinct(episode_name)

# Doesn't look as great when I try this
lines %>% 
  filter(imdb_rating >= 9.0) %>% 
  count(season)

# You can also save the results of a pipeline. Here I want to get the top 3 directors by number of lines. I can chain together a pipeline to return this and save it to a new dataframe called top_dirs
top_dirs <- lines %>% 
  count(director) %>% 
  arrange(desc(n)) %>% 
  head(3)

#Advanced Method
#top_dirs <- lines %>% 
#  count(director) %>% 
#  arrange(desc(n)) %>% 
#  head(3) %>% 
#  select(director) %>% 
#  unlist()
```

### the %in% operator

The %in% operator is a useful tool when you have a series of conditions that you want to check. Instead of asking whether cond1 or cond2 or cond3 .... are met, you can use the %in% operator to check them all at once.

```{r}
tops <- c("Paul Fieg", "Randall Einhorn", "Greg Daniels")

#The next line of code will look through the tops vector and will check if each of the values in the vector are found in the director column
lines %>% 
  filter(director %in% tops)

#Advanced Method - Skipping the creating a vector step.
#top_dirs <- lines %>% 
#  count(director) %>% 
#  arrange(desc(n)) %>% 
#  head(3) %>% 
#  select(director) %>% 
#  unlist()

# lines %>% 
#   filter(director %in% top_dirs)

```

## select()

Whereas the `filter()` function allows you to return only certain _rows_ matching a condition, the `select()` function returns only certain _columns_. The first argument is the data, and subsequent arguments are the columns you want.

```{r}
# You can simply list the column names you want to keep separated by a column
lines %>% 
  select(episode_name, director)

# Or you can use the - sign to tell dplyr the columns you want to drop
lines %>% 
  select(-director, -season)
```

### starts_with(), ends_with(), contains()

The `starts_with`, `ends_with` and `contains` functions provide very useful tools for dropping/keeping several variables at once without having to list each and every column you want to keep. The function will return columns that either start with a specific string of text, ends with a certain string of text, or contain a certain string of text.

```{r}
# As a note, these functions are all case sensitive
lines %>% 
  select(starts_with("text"))

lines %>% 
  select(-ends_with("r"))

lines %>% 
  select(contains("rec"))

# You can also create a vector of column names you want to keep 
lines %>% 
  names()

vars <- c("index", "season","episode", "episode_name")

lines %>% 
  select(vars)

```

However, you can go ahead and combine `filter` and `select` to subset your data even further.

```{r}
lines %>% 
  filter(imdb_rating >= 9.0) %>% 
  select(vars)

lines %>% 
  filter(director %in% top_dirs) %>% 
  select(vars)
```

## Exercise 

Using the `lines` dataset

1. Using the `lines` dataset, how many directors made Office episodes with an imdb rating less than 7  Answer should be 2. `hint: use filter and distinct`

2. Using the `office_ratings` dataset, use filter to find the episodes with an imdb rating between 7 and 7.5, followed by select to show the title, imdb rating, and air date for these observations, arranged in ascending order by air date.

3. If I tried to arrange the answer from question 2 by total votes instead of air date, would that work? Why or why not? 

<details><summary>Click here for the answers </summary>

```{r}
#1. 
lines %>% 
  filter(imdb_rating < 7) %>% 
  distinct(director)

#2.
office_ratings %>% 
  filter(imdb_rating >= 7 & imdb_rating <= 7.5) %>% 
  select(title, imdb_rating, air_date) %>% 
  arrange(air_date)

#3. 
#It Would not work because total_votes is no longer a part of the dataset as we selected title, imdb_rating, and air_date

```

</details>

-------------------------

## summarize()

The `summarize()` function summarizes multiple values to a single value. On its own the `summarize()` function doesn't seem to be all that useful. The dplyr package provides a few convenience functions called `n()` and `n_distinct()` that tell you the number of observations or the number of distinct values of a particular variable.

Notice that summarize takes a data frame and returns a data frame. In this case it's a 1x1 data frame with a single row and a single column. The name of the column, by default is whatever the expression was used to summarize the data. This usually isn't pretty, and if we wanted to work with this resulting data frame later on, we'd want to name that returned value something easier to deal with.

```{r}
lines %>% 
  summarize(mean(imdb_rating))

#Gotta watch out for nas. Use na.rm = TRUE to run the calculation without nas being involved.
lines %>% 
  summarize(mean(imdb_rating, na.rm = TRUE))

#The default column names do not look very aesthetically pleasing, let's take care of that.
lines %>% 
  summarize(mean_rating = mean(imdb_rating, na.rm = TRUE))

#Again, summarize alone is not entirely useful
lines %>% 
  summarize(n())

```

## group_by()

We saw that `summarize()` isn't that useful on its own. Neither is `group_by()` All this does is takes an existing data frame and coverts it into a grouped data frame where operations are performed by group.

```{r}
lines %>% 
  group_by(director)

lines %>% 
  group_by(director, season)
```

The real power comes in where `group_by()` and `summarize()` are used together. First, write the `group_by()` statement. Then pipe the result to a call to `summarize()`.

```{r}
#Let's summarize the mean imdb_rating for each director
lines %>% 
  group_by(director) %>% 
  summarize(mean_rating = mean(imdb_rating, na.rm = TRUE))
  
lines %>% 
  group_by(director) %>% 
  summarize(mean_rating = mean(imdb_rating, na.rm = TRUE)) %>% 
  arrange(desc(mean_rating))

#Let's see how many episodes each director directed
lines %>% 
  group_by(director) %>% 
  summarize(n_episodes = n_distinct(episode)) %>% 
  arrange(desc(n_episodes)) 

#Let's calculate who had the most lines
lines %>% 
  group_by(character) %>% 
  summarize(n_lines = n()) %>% 
  arrange(desc(n_lines))

lines %>% 
  group_by(character) %>% 
  summarize(n_lines = n()) %>% 
  arrange(desc(n_lines)) %>% 
  head(4)

#Taking it deeper, we can try and calculate who had the most lines in each episode. In this we use the top_n function, which will be taking only the observation from each episode that has the highest n (number of lines)
top_lines <- lines %>% 
  group_by(episode_name, character) %>% 
  summarize(n_lines = n(),
            mean_rating = mean(imdb_rating, na.rm = TRUE)) %>% 
  arrange(episode_name, desc(n_lines)) %>% 
  top_n(n = 1, wt=n_lines) 

top_lines

#You can save your pipeline and then use that to do even more analysis/investigation
  top_lines %>% 
    group_by(character) %>% 
    summarize(mean = mean(mean_rating, na.rm = TRUE), n_lines = n()) %>% 
    arrange(desc(mean))

```

## Exercise 

1. Using the `top_lines` dataset, show the episode_name, mean_rating, and number of lines spoken for when Erin is the character with the most lines. Sort by descending mean_rating _Hint:_ 3 pipes: `filter`, `select`, and `arrange`.


2. Using the `office_ratings` dataset, determine which season had the highest mean imdb rating. Include the standard deviation (`sd()`) for each season. _Hint:_ 3 pipes: `group_by`, `summarize`, and `arrange`.

<details><summary>Click here for the answers </summary>
```{r}
#1. 
top_lines %>% 
  filter(character == "Erin") %>% 
  select(episode_name, mean_rating, n_lines) %>% 
  arrange(desc(mean_rating))

#2. 
office_ratings %>% 
  group_by(season) %>% 
  summarize(mean_rating = mean(imdb_rating, na.rm = TRUE),
            sd_rating = sd(imdb_rating, na.rm = TRUE)) %>% 
  arrange(desc(mean_rating))

```

</details>
-------------------------

## Writing out data

Just like you can read in data from read_csv(), you can also write out a csv file. Let's create a new file for our `top_lines` data.

```{r}
#Include the name of the object to export (top_lines) and the name of the file
write_csv(top_lines, "top_lines.csv")
```