---
title: "Data Visualization with ggplot2 - Exercises"
output:
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

# Set up

```{r}
# Load up the tidyverse
library(tidyverse)
```

# Read in data

We are going to use a dataset comparing food consumption and co2 emmissions. Use the `read_csv` package to read in the data.

```{r}
# We are going to save the food.csv file as "food"
food <- read_csv('food.csv') 

# For these exercises, we are going to look at everything that is not meat or fish
other_food <- c("Eggs", "Milk - inc. cheese", "Wheat and Wheat Products", "Rice", "Soybeans", "Nuts inc. Peanut Butter")

# We are going to filter so that we only have the other_food.
food <- food %>% 
  filter(food_category %in% other_food)
```

# Exercise 1

1. Create a single scatter plot investigating the relationship between `consumption` and `co2_emmission`. 

2. Then, create a scatter plot that has colors and shapes determined by `food_category`. Facet this plot by `continent`

3. Provide appropriate labels/titles to this plot. 

4. Add the theme_classic() theme to your plot.

<details><summary>Click here for the answers </summary>

```{r}
#1.
ggplot(food, aes(x = co2_emmission, y = consumption)) +
  geom_point()

#2.
# You could either put the aes call to shape and color in the geom or in the ggplot() function itself
p <- ggplot(food, aes(x = co2_emmission, y = consumption, color = food_category, shape = food_category)) + 
  geom_point() + 
  facet_wrap(~continent)

#3.
# Notice here that I have to add in a portion for both the color and the shape.
p <- p + labs(x = "CO2 Emmissions (kg co2/person/day)",
         y = "Consumption (kg/person/day)",
         title = "Consumption vs. CO2 Emmissions",
         subtitle = "Non-meat and fish",
         color = "Food Category", 
         shape = "Food Category")
#4.
p + theme_classic()
```
</details>

# Exercise 2 

Start by running this code:
```{r}
food_sum <- food %>% 
  filter(continent != "Oceania") %>% 
  group_by(country, continent) %>% 
  summarize(co2_emmission = sum(co2_emmission),
            consumption = sum(consumption)) %>% 
  mutate(co2_per_consumption = co2_emmission/consumption)
```

And then recreate the following plot:


```{r, echo = FALSE}
ggplot(food_sum, aes(x = reorder(continent, desc(co2_per_consumption)), y = co2_per_consumption, fill = continent, color = continent)) + 
  geom_violin() + 
  geom_jitter() + 
  labs(x = "Continent", 
       y = "Emmissions per Consumption",
       fill = "Continent", 
       color = "Continent", 
       title = "Emmissions per Consumption by Continent",
       subtitle = "Ordered by Emmissions per Consumption") +
  theme_classic()
```

<details><summary>Click here for the code </summary>

```{r, fig.show = "hide"}
ggplot(food_sum, aes(x = reorder(continent, desc(co2_per_consumption)), y = co2_per_consumption, fill = continent, color = continent)) + 
  geom_violin() + 
  geom_jitter() + 
  labs(x = "Continent", 
       y = "Emmissions per Consumption",
       fill = "Continent", 
       color = "Continent", 
       title = "Emmissions per Consumption by Continent",
       subtitle = "Ordered by Emmissions per Consumption") +
  theme_classic()
```
</details>

