---
title: "Untitled"
author: "David Martin"
date: "February 7, 2019"
output: 
  html_document: 
    toc: yes  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This is my first RMarkdown document!

# Let's embed some R code

Let's load the **Gapminder** data:

```{r}
gm <- read.csv("Data/gapminder.csv")
head(gm)
summary(gm$lifeExp)
```

The mean life expectancy is `r mean(gm$lifeExp)` years.

The years surveyed in this data include: `r unique(gm$year)`.

### Chunk options

You can modify the behavior of an R chunk with [options](http://yihui.name/knitr/options/). Options are passed in after a comma on the fence, as shown below. 

```{r optionalChunkName, echo=TRUE, results='hide'}
# R code here
```

Some commonly used options include:

- `echo`: (`TRUE` by default) whether to include R source code in the output file.
- `results` takes several possible values:
    - `markup` (the default) takes the result of the R evaluation and turns it into markdown that is rendered as usual.
    - `hide` will hide results.
    - `hold` will hold all the output pieces and push them to the end of a chunk. Useful if you're running commands that result in lots of little pieces of output in the same chunk.
    - `asis` writes the raw results from R directly into the document. Only really useful for tables.
- `include`: (`TRUE` by default) if this is set to `FALSE` the R code is still evaluated, but neither the code nor the results are returned in the output document. 
- `fig.width`, `fig.height`: used to control the size of graphics in the output.

Try modifying your first R chunk to use different values for `echo`, `results`, and `include`.
