---
title: 'R: An Introduction'
output: pdf_document
---

```{r, echo=FALSE, message=FALSE, eval=TRUE}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, results="hide", cache=TRUE)
options(digits=3)
options(max.print=200)
.ex <- 1 # Track ex numbers w/ hidden var. Increment each ex: `r .ex``r .ex=.ex+1`
```

##Homework 1


###EXERCISE `r .ex``r .ex=.ex+1`

1.	You have a patient with a height (inches) of 73 and a weight (lbs) of 203. Create r objects labeled 'height' and 'weight'.

```{r}
height <- 73
weight <- 203
```

2.	Convert 'weight' to 'weight_kg' by dividing by 2.2. Convert 'height' to 'height_m' by dividing by 39.37
```{r}
weight_kg <- weight /2.2
height_m <- height/39.37
```

3.	Calculate a new object 'bmi' where BMI = weight_kg / (height_m*height_m)
```{r}
bmi <- weight_kg / (height_m * height_m)
bmi <- weight_kg / (height_m ^ 2)
```


#------------------------------------------------
###EXERCISE `r .ex``r .ex=.ex+1`

See `?abs` and calculate the square root of the log-base-10 of the absolute value of `-4*(2550-50)`. Answer should be 
`2`.
```{r}
sqrt(log(abs(-4*(2550 - 50)), 10))
```


#------------------------------------------------------
The data we're going to look at is cleaned up version of a gene expression dataset from [Brauer et al. Coordination of Growth Rate, Cell Cycle, Stress Response, and Metabolic Activity in Yeast (2008) _Mol Biol Cell_ 19:352-367](http://www.ncbi.nlm.nih.gov/pubmed/17959824). This data is from a gene expression microarray, and in this paper the authors are examining the relationship between growth rate and gene expression in yeast cultures limited by one of six different nutrients (glucose, leucine, ammonium, sulfate, phosphate, uracil). If you give yeast a rich media loaded with nutrients except restrict the supply of a _single_ nutrient, you can control the growth rate to any rate you choose. By starving yeast of specific nutrients you can find genes that: 

1. **Raise or lower their expression in response to growth rate**. Growth-rate dependent expression patterns can tell us a lot about cell cycle control, and how the cell responds to stress. The authors found that expression of >25% of all yeast genes is linearly correlated with growth rate, independent of the limiting nutrient. They also found that the subset of negatively growth-correlated genes is enriched for peroxisomal functions, and positively correlated genes mainly encode ribosomal functions. 
2. **Respond differently when different nutrients are being limited**. If you see particular genes that respond very differently when a nutrient is sharply restricted, these genes might be involved in the transport or metabolism of that specific nutrient.



###EXERCISE `r .ex``r .ex=.ex+1`

1. What's the standard deviation of the life expectancy (hint: get help on the `sd` function with `?sd`).
```{r}
 sd(gm$lifeExp, na.rm = TRUE)
```

2. What's the mean population size in millions? (hint: divide by `1000000`, or alternatively, `1e6`).
```{r}
 mean(gm$pop, na.rm = TRUE) / 1e6
```

3. What's the range of years represented in the data? (hint: `range()`).
```{r}
 range(gm$year)
 range(gm$pop)
```
4.	Run a correlation between life expectancy and GDP per capita (hint: ?cor())
```{r}
 cor(gm$lifeExp, gm$gdpPercap, method = "kendall")
```
